Investigating Biases Associated with Dietary Starch Incorporation into an Oral Biofilm Model

### 2021-09-20 analysis/paper/paper.Rmd
revisions to first draft

---
title: "Untitled"
author:
  - "Bjørn Peare Bartholdy"
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
  bookdown::html_document2:
    df_print: kable
    keep_md: true
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
keywords: |
  keyword 1; keyword 2
---

@@

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus, mineralised dental plaque, has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculusCooking2014]. 
It has also proven to be able to retain and preserve dietary information for millenea,
providing detailed dietary information on past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, few studies directly investigated the presence of plant microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. The ability to detect coca
consumption by looking at human dental calculus was explored by Klepinger and Kuhn
[-@klepingerCalculusCoca1977], and starches and phytoliths were extracted from 
human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].

In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has been, 
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- clearly stating the general problem being addressed by this particular study -->
<!-- Dental calculus research has its limitations, all of which I will solve with this
single method. -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways and the methods we use to 
extract and analyse dental calculus and make inferences on past diets. 
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. Other studies have been conducted on modern populations 
[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
well-documented dietary intake, to reconcile our interpretations on the recovery 
of microremains (both phytoliths and starches) from dental calculus and how well
they represent actual dietary intake. These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the rate of $\alpha$-amylase differs both between and within
individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and you have a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus in a controlled laboratory 
setting, allowing us full control over dietary input. We use a multiwell setup for consistent, high-throughout analysis.
Our main questions concern the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of 
diet from extracted microremains?  
We find that, despite an absence of $\alpha$-amylase in
the model, a limited proportion of the starch input is actually 
retained in the calculus, as well as a shift in the size ratios of individual 
starch granules
that are incoporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit 
increases.

@@

```{r protocol-fig, fig.cap="Overview of experiment protocol"}
knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
```
<!-- modify sampling days from protocol picture to match amylase sampling -->

@@

@@

<!-- add to protocols.io -->
Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
with no history of caries, who abstained from oral hygiene for 24 hours, and no 
food was consumed two hours prior to donation. No antibiotics were taken up to six 
months prior to donation. 
The saliva was filtered through a sterilised (with bleach) nylon cloth to remove
particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
at 36&deg;C, 30 rpm. The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system before
starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
(w/v) sucrose solution. Before transferring biofilm samples to the starch treatment plate, 
the plates were agitated to keep the starches in suspension in the solutions. 
During treatments, the rpm was increased to 60 to facilitate contact between
starch granules and biofilms.

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a... 
<!-- photometric reader specs -->
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
For a more detailed protocol, see <!-- protocols.io DOI -->

Starch granule counts were divided into three size categories: small (<10 $\mu$m),
medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
potato starches were combined for the mixed-treatment samples, as it was not 
possible to distinguish betweem the small starch granules from the two species.

### Treatment solutions

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
counted (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts were extrapolated 
(see Supplementary material for more details).

### Extraction method

Extraction of starches from the calculus samples was performed by dissolving the
calculus in ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. Wheat and mix counts were extrapolated as described above.

@@

Statistical analysis was conducted in `r version$version.string`  [@R-base].
<!-- R packages used -->
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted to explore the effect of the different 
starch treatments on biofilm growth.

@@

count_sol_out <- count_sol %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
  ungroup() %>%
  add_row(tibble_row(count_sol_comb_out), .before = 3)
```

```{r solution-count-tbl}
cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
knitr::kable(count_sol_out, col.names = cnames, 
             caption = "Mean starch counts from solutions, including the
             proportional makeup of the different sizes of granules.")
```

@@

count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
  select(1,2,3,7,4,8,5,9,6,10) #%>%
  #add_row(tibble_row(count_sample_comb_out), .before = 3)
```

@@

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m,
small potato starches were counted as wheat starches in the mixed-treatment samples. This is reasonable given that the small potato granules make up an insignificant proportion of the 
total count of small granules within mixed-treatment solutions, which are predominantly
wheat granules
(`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`).
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = `r subset(count_sol, solution == "mix")$total`),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules 
(mean = `r subset(count_sample, treatment == "wheat")$total`). 
The potato treatment had the lowest absolute counts in both the solution 
(`r subset(count_sol, solution == "potato")$total`)
and in the biofilm samples 
(`r subset(count_sample, treatment == "potato")$total`) 
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

<!-- look at proportions of incorporated starches in each sample before looking at overall means -->

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likeliness of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc$total), max(perc$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(`r format(count_sol$total[4], big.mark = ",")`) 
and potato
(`r format(count_sol$total[3], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment; although, starch availability is unlikely
to be a main factor in starch incorporation. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc$total), max(perc$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010].

@@

<!-- although this study extrapolated counts to full sample content, which is
rarely done in archaeological studies-->

<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules, and plants that 
produce predominantly larger starch (10-20 $\mu$m) granules were over-represented.
Both studies concluded that the dietary breadth of a population can
be captured in a sufficiently large sample, but interpretations on individual diets
from microremains in dental calculus is tenuous, at best 
[@leonardDentalCalculus2015; @powerChimpCalculus2015].

<!-- secondary findings -->

>>>>

---
title: "Investigating Biases Associated with Dietary Starch Incorporation into an Oral Biofilm Model"
author:
  - Bjørn Peare Bartholdy
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

@@

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus, mineralised dental plaque, has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculusCooking2014]. 
This dietary information can be preserved within calculus over many millenia,
providing a unique window into the food-related behaviours of past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has been, 
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored the how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and you have a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus in a controlled laboratory 
setting, allowing us full control over dietary input. Our main questions concern
the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of 
diet from extracted microremains?  
We find that, despite the absence of $\alpha$-amylase in
the model, a limited proportion of the starch input is actually 
retained in the calculus, as well as a shift in the size ratios of individual 
starch granules
that are incorporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit 
increases.

@@

@@

<!-- insert figure of the setup -->

@@

```{r protocol-fig, fig.cap="Overview of experiment protocol"}
knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
```

<!-- modify sampling days from protocol picture to match amylase sampling -->

Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
with no history of caries, who abstained from oral hygiene for 24 hours, and no 
food was consumed two hours prior to donation. No antibiotics were taken up to six 
months prior to donation. 
The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
nylon cloth to remove particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis from saliva inoculation days. 
An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system before
starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch treatments were created in dH~2~O
with 5% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
plate, the plates were agitated to keep the starches in suspension in the
solutions. During treatments, the rpm was increased to 60 to facilitate contact
between starch granules and biofilms.

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
For a more detailed protocol, see <!-- protocols.io DOI -->

## Treatment solutions

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
were extrapolated (see Supplementary material for more details). For potato
treatment samples, the whole slide was counted.

## Extraction method

Extraction of starches from the calculus samples was performed by dissolving the
calculus in -0.5 M<!-- what percentage???--> ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. Wheat and mix counts were extrapolated as described above.
Both wheat and potato granules were divided into three size categories: 
small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).

@@

Statistical analysis was conducted in `r version$version.string`  [@R-base] and
the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
here [@R-here], and patchwork [@R-patchwork].
<!-- R packages used -->
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted to explore the effect of the different 
starch treatments on biofilm growth.

@@

count_sol_out <- count_sol %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
  ungroup() %>%
  add_row(tibble_row(count_sol_comb_out), .before = 3)

count_sol_out[1,3] <- NA
```

@@

count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
  select(1,2,3,7,4,8,5,9,6,10) #%>%

count_samp_out[1,3] <- NA # replace with NA
  #add_row(tibble_row(count_sample_comb_out), .before = 3)
```

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. This is a
reasonable assumption to make based on the small granule counts from the separate
treatment solutions, where the wheat granules made up 
`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`
of all small granule counts in the separated solutions.
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

```{r solution-count-tbl}
cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
options(knitr.kable.NA = "")
knitr::kable(count_sol_out, col.names = cnames, 
             caption = "Mean starch counts from solutions, including the
             proportional makeup of the different sizes of granules.")
```

@@

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(subset(count_sol, solution == "mix")$total, 3)`),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules 
(mean = `r signif(subset(count_sample, treatment == "wheat")$total, 3)`). 
The potato treatment had the lowest absolute counts in both the solution 
(`r signif(subset(count_sol, solution == "potato")$total, 3)`)
and in the biofilm samples 
(`r signif(subset(count_sample, treatment == "potato")$total, 3)`) 
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc$total), max(perc$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(count_sol$total[4], big.mark = ",")`) 
and potato
(mean = `r format(count_sol$total[3], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc$total), max(perc$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010]. <!-- although this study extrapolated counts to full sample content, which is
rarely done in archaeological studies -->This could, in part, be due to the 
activity of oral amylase, which was absent in our model.

@@

<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.

<!-- secondary findings -->
We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10 -- 20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
This was especially true for the potato starches, which can reach up to 100 $\mu$m
in diameter, whereas wheat generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
The size bias is consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
which found that plants producing starches 10--20 $\mu$m in
size were over-represented; although, the representation of granules larger than
20 $\mu$m is unclear.  
Given the differences in proportional incorporation between wheat and potato,
especially for small and medium granules, it is also possible that the morphology
of granules plays a role. However, the proportional count differences between large 
wheat and potato granules were less pronounced.

<!-- Since those results were obtained -->
<!-- on a modern population of chimps that were only buried for a short period of time,  -->
<!-- degradation was likely limited [@powerChimpCalculus2015]. -->

The taphonomy of starch granules has been addressed previously, and depends on 
burial environment (pH, temperature, moisture content, microorganisms), as well 
as processing of the granules prior to deposition 
[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
The potential for detrimental diagenetic effects on starch granules increases as 
the size and amylose content decreases
[@francoStarchDegradation1992; @haslamDecompositionStarch2004].

<!-- add potential for morphological differences affecting incorporation 
 Medium potato starches had the highest rate of incorporation 
 All large starches had a consistently low rate of incorporation 
 Starch granule adhesion to bacteria and enzymes? Do they have a preference?
 Binding to select species in a biofilm could explain clustering
 Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017-->

<!-- Combined, the effect of intra-oral starch incorporation -->
<!-- and the bias against large granules, together with the increased effect of taphonomy -->
<!-- on  small granules, would eliminate a large portion of the consumed starch granules, -->
<!-- and could explain the large descrepency between our counts (unaffected by hydrolysis  -->
<!-- and taphonomy) and the aforementioned studies.  -->

Another potentially important factor 
from our results is the size of the calculus deposit, which seems to influence the 
quantity of starches extracted from the calculus, as we found a strong positive 
correlation between biofilm size and retained starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2));
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
The lack of a correlation
in archaeological contexts could be due to diagenetic effects. Alternately, the absence of 
amylase activity in our study could be impacting our interpretations.

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? - adhesion to bacteria
- stuck in cracks and crevices
- covered by new layer of plaque
